Determining the direction of causality in the face of measurement error

### 2016-02-16 docs/manuscript.Rmd
fixed missing pic

---
title: "Determining the direction of causality in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: true
bibliography: manuscript.bib
csl: nature.csl
---

@@

The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is highly liable to measurement error.

Limitations

Pleiotropy

$$
x = \alpha_g + \beta_g g + \epsilon_g \\
x_O = \alpha_m + \beta_m x + \epsilon_m \\
y = \alpha_x + \beta_x x + \epsilon_x
$$

$$
\begin{aligned}
cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                & = D\beta_x\beta_g var(g)
\end{aligned}
$$

$$
D = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
$$

## Tables

@@

## Supplementary tables

@@

## Supplementary figures

>>>>

---
title: "Determining the direction of causality in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  word_document
bibliography: manuscript.bib
csl: nature.csl
---

@@

The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is very susceptible to measurement error. With the presence of measurement error near ubiquitous in most cross sectional studies, and our ability to measure it limited, we argue that it should be given serious consideration in designing statistical analyses that attempt to make causal inference, and that any putitive results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.

@@

$$
\begin{aligned}
x & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_m + \beta_m x + \epsilon_m \\
y & = \alpha_x + \beta_x x + \epsilon_x
\end{aligned}
$$

$$
\begin{aligned}
cov(g, \hat{y}) & = \frac{\beta_x\beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_x\beta_g var(g)
\end{aligned}
$$

@@

### 2016-02-16 docs/manuscript.tex
fixed missing pic

\setlength{\droptitle}{-2em}
  \title{Determining the direction of causality in the face of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{14 February 2016}

@@

Figure 3: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
is negative we expect the Steiger test to be more likely to be wrong
about the direction of causality. Rows of graphs represent the sample
size used in the simulations.

>>>>

\setlength{\droptitle}{-2em}
  \title{Determining the direction of causality in the face of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{16 February 2016}

@@

\includegraphics{manuscript_files/figure-latex/cit_mr_comparison_figure-1.pdf}\\
Figure 3: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
is negative we expect the Steiger test to be more likely to be wrong
about the direction of causality. Rows of graphs represent the sample
size used in the simulations.
